Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-07 talk/icooolps2009/paper.tex
use proper way of doing "

In this paper we discuss ongoing work in the PyPy project to improve the
performance of interpreters written with the help of the PyPy toolchain. The
approach is that of a tracing JIT compiler. Contrary to the tracing JITs for dynamic
languages that currently exist, PyPy's tracing JIT operates "one level down",
e.g. it traces the execution of the interpreter, as opposed to the execution
of the user program. The fact that the program the tracing JIT compiles is
in our case always an interpreter brings its own set of problems. We describe
tracing JITs and their application to interpreters in Section
\ref{sect:tracing}.  By this approach we hope to arrive at a JIT compiler that can be
applied to a variety of dynamic languages, given an appropriate interpreter for each of them. The
process is not completely automatic but needs a small number of hints from the
interpreter author, to help the tracing JIT. The details of how the process
integrates into the rest of PyPy will be explained in Section
\ref{sect:implementation}. This work is not finished, but has already produced some
promising results, which we will discuss in Section \ref{sect:evaluation}.

@@

%- dynamic languages important
%- notoriously difficult to achieve good performance
%- even though the techniques exist since a while, not many implementations
%  actually use them
%    - hard to get all corner-cases right
%    - languages evolve
%    - modern dynamic languages are large
%    - open source/research communities don't have that many resources
%
%- PyPy project: trying find approaches to ease the implementation of dynamic
%languages
%- explore general ways to improve on the speed of dynamic languages with reduced
%effort
%- approach: write a tracing JIT that is applicable to many different languages,
%by tracing "one level done"
%- needs some hints by the interpreter-writer + slightly different optimizations
%- paper will describe the problems of applying a tracing jit to an interpreter
%- different integration needed than a typical tracing jit

@@

The PyPy project\footnote{http://codespeak.net/pypy}
\cite{rigo_pypys_2006,carl_friedrich_bolz_to_2007} was started to implement a
new Python interpreter in Python but has now extended its goals to be an
environment where flexible implementation of dynamic languages can be written.
To implement a dynamic language with PyPy, an interpreter for that language has
to be written in RPython \cite{ancona_rpython:step_2007}. RPython ("Restricted Python") is a subset of Python
chosen in such a way that type inference can be performed on it. The language
interpreter can then be translated with the help of PyPy into various target
environments, such as C/Posix, the CLI and the JVM. This is done by a component
of PyPy called the \emph{translation toolchain}.

@@

This process assumes that the path through the loop that was traced is a
"typical" example of possible paths (which is statistically likely). Of course
it is possible that later another path through the loop is taken, in which case
one of the guards that were put into the machine code will fail. There are more
complex mechanisms in place to still produce more code for the cases of guard
failures \cite{XXX}, but they are independent of the issues discussed in this
paper.

@@

The tracing JIT of the PyPy project is atypical in that it is not applied to the
user program, but to the interpreter running the user program. In this section
we will explore what problems this brings, and suggest how to solve them (at least
partially). This means that there are two interpreters involved, and we need appropriate
terminology to distinguish beween them. On the one hand, there is the interpreter that
the tracing JIT uses to perform tracing. This we will call the \emph{tracing
interpreter}. On the other hand, there is the interpreter that runs the
user's programs, which we will call the \emph{language interpreter}. In the
following, we will assume that the language interpreter is bytecode-based. The
program that the language interpreter executes we will call the \emph{user
program} (from the point of view of a VM author, the "user" is a programmer
using the VM).

@@

In addition to the classification of the variables, there are two methods of
\texttt{JitDriver} that need to be called. Both of them receive as arguments the
current values of the variables listed in the definition of the driver. The
first one is \texttt{jit\_merge\_point} which needs to be put at the beginning
of the body of the bytecode dispatch loop. The other, more interesting one, is
\texttt{can\_enter\_jit}. This method needs to be called at the end of any
instruction that can set the program counter of the language interpreter to an
earlier value. For the example this is only the \texttt{JUMP\_IF\_A}
instruction, and only if it is actually a backward jump. The place where this
method is called is where the language interpreter performs profiling to decide
when to start tracing. It is also the place where the tracing JIT checks
whether a loop is closed. This is considered to be the case when the values of
the "green" variables are the same as at an earlier call to the
\texttt{can\_enter\_jit} method.

@@

The simple insight how to greatly improve the situation is that most of the
operations in the trace are actually concerned with manipulating the
bytecode and the program counter. Those are stored in variables that are part of
the position key (they are "green"), that means that the tracer checks that they
are some fixed value at the beginning of the loop. In the example the check
would be that the \texttt{bytecode} variable is the bytecode string
corresponding to the square function and that the \texttt{pc} variable is
\texttt{4}. Therefore it is possible to constant-fold computations on them away,
as long as the operations are side-effect free. Since strings are immutable in
RPython, it is possible to constant-fold the \texttt{strgetitem} operation. The
\texttt{int\_add} are additions of the green variable \texttt{pc} and a constant
number, so they can be folded away as well.

@@

When a hot user loop is identified, tracing is started. The
tracing interpreter is invoked to start tracing the language interpreter that is
running the user program. Of course the tracing interpreter cannot actually
trace the execution of the C representation of the language interpreter. Instead
it takes the state of the execution of the language interpreter and starts
tracing using a bytecode representation of the language interpreter. That means
there are two "versions" of the language interpreter embedded in the final
executable of the VM: on the one hand it is there as executable machine code, on
the other hand as bytecode for the tracing interpreter. It also means that
tracing is costly as it incurs a double interpretation overhead.

>>>>

In this paper we discuss ongoing work in the PyPy project to improve the
performance of interpreters written with the help of the PyPy toolchain. The
approach is that of a tracing JIT compiler. Contrary to the tracing JITs for dynamic
languages that currently exist, PyPy's tracing JIT operates ``one level down'',
e.g. it traces the execution of the interpreter, as opposed to the execution
of the user program. The fact that the program the tracing JIT compiles is
in our case always an interpreter brings its own set of problems. We describe
tracing JITs and their application to interpreters in Section
\ref{sect:tracing}.  By this approach we hope to arrive at a JIT compiler that can be
applied to a variety of dynamic languages, given an appropriate interpreter for each of them. The
process is not completely automatic but needs a small number of hints from the
interpreter author, to help the tracing JIT. The details of how the process
integrates into the rest of PyPy will be explained in Section
\ref{sect:implementation}. This work is not finished, but has already produced some
promising results, which we will discuss in Section \ref{sect:evaluation}.

@@

%- dynamic languages important
%- notoriously difficult to achieve good performance
%- even though the techniques exist since a while, not many implementations
%  actually use them
%    - hard to get all corner-cases right
%    - languages evolve
%    - modern dynamic languages are large
%    - open source/research communities don't have that many resources
%
%- PyPy project: trying find approaches to ease the implementation of dynamic
%languages
%- explore general ways to improve on the speed of dynamic languages with reduced
%effort
%- approach: write a tracing JIT that is applicable to many different languages,
%by tracing ``one level done''
%- needs some hints by the interpreter-writer + slightly different optimizations
%- paper will describe the problems of applying a tracing jit to an interpreter
%- different integration needed than a typical tracing jit

@@

The PyPy project\footnote{http://codespeak.net/pypy}
\cite{rigo_pypys_2006,carl_friedrich_bolz_to_2007} was started to implement a
new Python interpreter in Python but has now extended its goals to be an
environment where flexible implementation of dynamic languages can be written.
To implement a dynamic language with PyPy, an interpreter for that language has
to be written in RPython \cite{ancona_rpython:step_2007}. RPython (``Restricted
Python'') is a subset of Python
chosen in such a way that type inference can be performed on it. The language
interpreter can then be translated with the help of PyPy into various target
environments, such as C/Posix, the CLI and the JVM. This is done by a component
of PyPy called the \emph{translation toolchain}.

@@

This process assumes that the path through the loop that was traced is a
``typical'' example of possible paths (which is statistically likely). Of course
it is possible that later another path through the loop is taken, in which case
one of the guards that were put into the machine code will fail. There are more
complex mechanisms in place to still produce more code for the cases of guard
failures \cite{XXX}, but they are independent of the issues discussed in this
paper.

@@

The tracing JIT of the PyPy project is atypical in that it is not applied to the
user program, but to the interpreter running the user program. In this section
we will explore what problems this brings, and suggest how to solve them (at least
partially). This means that there are two interpreters involved, and we need appropriate
terminology to distinguish beween them. On the one hand, there is the interpreter that
the tracing JIT uses to perform tracing. This we will call the \emph{tracing
interpreter}. On the other hand, there is the interpreter that runs the
user's programs, which we will call the \emph{language interpreter}. In the
following, we will assume that the language interpreter is bytecode-based. The
program that the language interpreter executes we will call the \emph{user
program} (from the point of view of a VM author, the ``user'' is a programmer
using the VM).

@@

In addition to the classification of the variables, there are two methods of
\texttt{JitDriver} that need to be called. Both of them receive as arguments the
current values of the variables listed in the definition of the driver. The
first one is \texttt{jit\_merge\_point} which needs to be put at the beginning
of the body of the bytecode dispatch loop. The other, more interesting one, is
\texttt{can\_enter\_jit}. This method needs to be called at the end of any
instruction that can set the program counter of the language interpreter to an
earlier value. For the example this is only the \texttt{JUMP\_IF\_A}
instruction, and only if it is actually a backward jump. The place where this
method is called is where the language interpreter performs profiling to decide
when to start tracing. It is also the place where the tracing JIT checks
whether a loop is closed. This is considered to be the case when the values of
the ``green'' variables are the same as at an earlier call to the
\texttt{can\_enter\_jit} method.

@@

The simple insight how to greatly improve the situation is that most of the
operations in the trace are actually concerned with manipulating the
bytecode and the program counter. Those are stored in variables that are part of
the position key (they are ``green''), that means that the tracer checks that they
are some fixed value at the beginning of the loop. In the example the check
would be that the \texttt{bytecode} variable is the bytecode string
corresponding to the square function and that the \texttt{pc} variable is
\texttt{4}. Therefore it is possible to constant-fold computations on them away,
as long as the operations are side-effect free. Since strings are immutable in
RPython, it is possible to constant-fold the \texttt{strgetitem} operation. The
\texttt{int\_add} are additions of the green variable \texttt{pc} and a constant
number, so they can be folded away as well.

@@

When a hot user loop is identified, tracing is started. The
tracing interpreter is invoked to start tracing the language interpreter that is
running the user program. Of course the tracing interpreter cannot actually
trace the execution of the C representation of the language interpreter. Instead
it takes the state of the execution of the language interpreter and starts
tracing using a bytecode representation of the language interpreter. That means
there are two ``versions'' of the language interpreter embedded in the final
executable of the VM: on the one hand it is there as executable machine code, on
the other hand as bytecode for the tracing interpreter. It also means that
tracing is costly as it incurs a double interpretation overhead.
